---
title: "1 Introduction"
---

## Tidy script

Few useful tips how to handle your scripts tidy:

1.  **use projects**

-   create directories data, results, maps, backup etc.

2.  **libraries**

-   list all the libraries you need at the beginning of the script

```{r}
#| warning: false
library(tidyverse)
library(readxl)
```

3.  **remarks**

-   `ctrl + shift + C` will change the text to non-active (marked with hash tags #)

-   add notes to your scripts, you will be grateful later

4.  **separate scripts or sections**

-   `ctrl + shift + R` insert named section

-   use 4x`#`

-   `Alt + O` fold all

-   `Shift + Alt + O` unfold all

5.  **names of variables**

-   short and easy to handle, without spaces, strange symbols

-   rename strange names one by one

-   `%>% rename (ReleveNr = "Releve number")`

-   or change all difficult patterns at once

-   RegEx Regular expression

```{r}
#| eval: false
env <- env %>%
   rename_all(~ str_replace_all(., c(
    "\\." = "",     #remove dots in the variable names
    "\\m²" = "m2",
    "\\°" = "deg",
    "\\%" = "perc", # remove symbol % and change it to perc
    "\\(" = "",
    "\\)" = "",
    "\\/" = "",
    "\\?" = "",
    "\\s", "."))) #remove spaces
```

6.  **piping**

-   pipe binds individual steps into a sequence

-   `ctrl+shift+M` inserts a pipe `%>%` or `|>`

**extra reading**

<https://www.tidyverse.org/blog/2017/12/workflow-vs-script/>

<https://davidzeleny.net/wiki/doku.php/recol:clean_and_tidy_script>
